manuscript_figures

### 2024-08-01 scripts/manuscript_figures.Rmd
updated figure 5 plots and stats

#combine rugosity and kelp biomass
master.dat2 <- left_join(mean_rugosity, morph.filt)

# combine for final dataframe
master.dat <- left_join(master.dat1, master.dat2)

@@

#select necessary columns
all.params.data <- master.dat %>% dplyr::select("YEAR", "SITE", "COVER", "Fish.Biomass..g.m2.", "mean_rugosity", "average_kelp_biomass")

#rename various columns in data frame
names(all.params.data)[3] <- "Kelp Cover (%)"
names(all.params.data)[4] <- "Fish Biomass (g/cm2)"
names(all.params.data)[5] <- "Rugosity"
names(all.params.data)[6] <- "Kelp Biomass (g/cm2)"

#reorder columns to show biomass by eachother
all.params.data <- all.params.data[, c(1,2, 4, 6, 5, 3)]

@@

# Select the relevant columns
selected_columns <- all.params.data[, c("Fish Biomass (g/cm2)", "Kelp Biomass (g/cm2)", "Rugosity", "Kelp Cover (%)")]

@@

# Check for normality
shapiro.test(all.params.data$`Kelp Cover (%)`)
shapiro.test(all.params.data$`Fish Biomass (g/cm2)`) 
shapiro.test(all.params.data$Rugosity)
shapiro.test(all.params.data$`Kelp Biomass (g/cm2)`)

@@

# Multiple regression between fish biomass (g/m2), kelp biomass (g/m2), kelp cover, rugosity, temperature in ggplot 
# combine all dataframes
```{r}
# Rename columns to remove spaces and special characters
names(all.params.data) <- c("YEAR", "SITE", "Fish_Biomass", "Rugosity", "Kelp_Cover", "Kelp_Biomass")

@@

# Check for zero or negative values
zero_or_negative <- sapply(all.params.data[, c("Fish_Biomass", "Rugosity", "Kelp_Biomass", "Kelp_Cover")], function(x) sum(x <= 0))
zero_or_negative

# Remove rows with zero or negative values in relevant columns
all.params.data <- all.params.data[all.params.data$Fish_Biomass > 0 & 
                                   all.params.data$Rugosity > 0 & 
                                   all.params.data$Kelp_Biomass > 0 & 
                                   all.params.data$Kelp_Cover > 0, ]

# Fit a multiple regression model
fit1 <- lm(log(Fish_Biomass + 1.1) ~ log(Rugosity + 1.1) * log(Kelp_Biomass + 1.1) * log(Kelp_Cover + 1.1), data = all.params.data)
summary(fit1)

# Diagnostic plots for the regression model
par(mfrow=c(2,2))
plot(fit1)

# Produce added multivariable plots
av_plots <- avPlots(fit1)

#The x-axis displays a single predictor variable and the y-axis displays the response variable.
#The blue line shows the association between the predictor variable and the response variable, while holding the value of all other predictor variables constant.
#The points that are labelled in each plot represent the 2 observations with the largest residuals and the 2 observations with the largest partial leverage.

#view results of model
summary_text1 <- capture.output(summary(fit1))
writeLines(summary_text1, "output/model.result.tables/rugosity.included/summary_original.model.txt")

@@

# Try another type of multiple regession plot
```{r}
#fit multiple linear regression model WITH SITE AND YEAR
fit2 <- lm(Fish_Biomass ~ SITE*YEAR, data = all.params.data)
summary(fit2)

#view results of model
summary_text2 <- capture.output(summary(fit2))
writeLines(summary_text2, "output/model.result.tables/rugosity.included/summary_model_site_year_fixed_fishbiomass.txt")

# Fit the mixed-effects model
fit3 <- lmer(log(Fish_Biomass) ~ log(Rugosity) * log(Kelp_Biomass) * log(Kelp_Cover) + 
              (1 | SITE) + (1 | YEAR), data = all.params.data)
summary(fit3)

#view results of model
summary_text3 <- capture.output(summary(fit3))
writeLines(summary_text3, "output/model.result.tables/rugosity.included/summary_model_site_year_random.txt")

# run with just site as random effect
# Fit the mixed-effects model
fit4 <- lmer(log(Fish_Biomass + 1.1) ~ log(Rugosity + 1.1) * log(Kelp_Biomass + 1.1) * log(Kelp_Cover + 1.1) + 
              (1 | SITE), data = all.params.data)
summary(fit4)

#view results of model
summary_text4 <- capture.output(summary(fit4))
writeLines(summary_text4, "output/model.result.tables/rugosity.included/summary_model_site_random.txt")

# site and year no significant main effects

# run with just year as random effect
# Fit the mixed-effects model
fit5 <- lmer(log(Fish_Biomass + 1.1) ~ log(Rugosity + 1.1) * log(Kelp_Biomass + 1.1) * log(Kelp_Cover + 1.1) + 
              (1 | YEAR), data = all.params.data)
summary(fit5)

#view results of model
summary_text5 <- capture.output(summary(fit5))
writeLines(summary_text5, "output/model.result.tables/rugosity.included/summary_model_year_random.txt")

#year is significantly different for kelp cover

# Run all above models without rugosity
```{r}

@@

# Now you can use this new 'Species' column in your ggplot
production_plot <- ggplot(pb_summary, aes(x = YEAR, y = mean_production, color = TREATMENT, group = TREATMENT)) +
  geom_point(size = 2) +         # Add points
  geom_line() +          # Add lines
  geom_errorbar(aes(ymin = mean_production - se_production, ymax = mean_production + se_production), width = 0.2) +  # Add error bars
  labs(x = "Year",  # X-axis label
       y = expression("Fish Production" ~ (g ~ m^-2 ~ yr^-1)),  # Y-axis label
       color = "Habitat") +  # Legend title
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 18),  # Rotate x-axis labels if needed
        axis.title.x = element_text(vjust = -0.5, size = 18),  # Adjust x-axis title position
        axis.title.y = element_text(vjust = 1.5, size = 18),
        axis.text.y = element_text(size = 18),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18)) +
  scale_y_continuous(limits = c(0, 8), expand = c(0, 0));production_plot # Additional attempt to change legend title  # Use viridis colorblind-friendly palette

# Save the plot
ggsave("output/figures/production.habitat.png", production_plot_A, width = 20, height = 10, units = "in", dpi = 300)

@@

# Calculate the differences
merged_data$production_difference <- merged_data$mean_production_kelp - merged_data$mean_production_control
merged_data$se_difference <- sqrt(merged_data$se_production_kelp^2 + merged_data$se_production_control^2)

# Step 4: Plot the data
production_diff_plot <- ggplot(merged_data, aes(x = YEAR, y = production_difference, group = TREATMENT_kelp)) +
  geom_point(size = 2) +  # Points with light blue outline
  geom_line() +  # Lines with light blue color
  geom_errorbar(aes(ymin = production_difference - se_difference, ymax = production_difference + se_difference), 
                width = 0.2, color = "black", size = 0.7) +  # Error bars
  labs(x = "Year",  # X-axis label
       y = expression("Kelp Habitat Fish Production" ~ (g ~ m^-2 ~ yr^-1)),  # Y-axis label
       color = "Treatment") +  # Legend title
  theme_bw() +  # White background with black lines
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 18),  # Rotate x-axis labels if needed
        axis.title.x = element_text(vjust = -0.5, size = 18),  # Adjust x-axis title position
        axis.title.y = element_text(vjust = 1.5, size = 18),  # Adjust y-axis title position
        axis.text.y = element_text(size = 18),  # Adjust y-axis text size
        legend.text = element_text(size = 16),  # Legend text size
        legend.title = element_text(size = 18));production_diff_plot

# Save the plot
ggsave("output/figures/kelp.fish.production.png", production_diff_plot, width = 20, height = 10, units = "in", dpi = 300)

@@

# Print the average
average_production_kelp # 4.359285 g/m2/yr
```

# Plot all two panels together using patchwork
```{r}
# Combine the plots with patchwork
prod_combined_plot <- (production_plot | production_diff_plot) +
  plot_annotation(tag_levels = 'A') + 
  theme(plot.tag = element_text(size = 20, face = "bold"));prod_combined_plot

>>>>

# combine for final dataframe
master.dat <- left_join(master.dat1, morph.filt)

@@

#select necessary columns
all.params.data <- master.dat %>% dplyr::select("YEAR", "SITE", "COVER", "Fish.Biomass..g.m2.", "average_kelp_biomass")

#rename various columns in data frame
names(all.params.data)[3] <- "Kelp Cover (%)"
names(all.params.data)[4] <- "Fish Biomass (g/cm2)"
names(all.params.data)[5] <- "Kelp Biomass (g/cm2)"

#reorder columns to show biomass by eachother
all.params.data <- all.params.data[, c(1,2, 4, 5, 3)]

@@

# Select the relevant columns
selected_columns <- all.params.data[, c("Fish Biomass (g/cm2)", "Kelp Biomass (g/cm2)", "Kelp Cover (%)")]

@@

# Check for normality
shapiro.test(all.params.data$`Kelp Cover (%)`)
shapiro.test(all.params.data$`Fish Biomass (g/cm2)`) 
shapiro.test(all.params.data$`Kelp Biomass (g/cm2)`)

@@

# Multiple regression between fish biomass (g/m2), kelp biomass (g/m2), kelp cover,  temperature in ggplot 
# combine all dataframes
```{r}
# Rename columns to remove spaces and special characters
names(all.params.data) <- c("YEAR", "SITE", "Fish_Biomass", "Kelp_Cover", "Kelp_Biomass")

@@

# Check for zero or negative values
zero_or_negative <- sapply(all.params.data[, c("Fish_Biomass", "Kelp_Biomass", "Kelp_Cover")], function(x) sum(x <= 0))
zero_or_negative

# Remove rows with zero or negative values in relevant columns
all.params.data <- all.params.data[all.params.data$Fish_Biomass > 0 & 
                                   all.params.data$Kelp_Biomass > 0 & 
                                   all.params.data$Kelp_Cover > 0, ]

@@

# Run all models without rugosity
```{r}

@@

# Now you can use this new 'Species' column in your ggplot
production_plot <- ggplot(pb_summary, aes(x = YEAR, y = mean_production, color = TREATMENT, group = TREATMENT)) +
  geom_point(size = 3) +         # Add points
  geom_line() +          # Add lines
  geom_errorbar(aes(ymin = mean_production - se_production, ymax = mean_production + se_production), width = 0.2) +  # Add error bars
  labs(x = "Year",  # X-axis label
       y = expression("Fish Production" ~ (g ~ m^-2 ~ yr^-1)),  # Y-axis label
       color = "Habitat") +  # Legend title
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
        axis.title.x = element_text(vjust = -0.5, size = 18),
        axis.title.y = element_text(vjust = 1.5, size = 18),
        axis.text.y = element_text(size = 18),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18),
        legend.position = "top") +
  scale_y_continuous(limits = c(0, 8), expand = c(0, 0));production_plot # Additional attempt to change legend title 
#blue color is #00BFC4

# Save the plot
ggsave("output/figures/production.habitat.png", production_plot, width = 20, height = 10, units = "in", dpi = 300)

@@

# Calculate the differences
merged_data$production_difference <- merged_data$mean_production_kelp + merged_data$mean_production_control
merged_data$se_difference <- sqrt(merged_data$se_production_kelp^2 + merged_data$se_production_control^2)

# Step 4: Plot the data
production_diff_plot <- ggplot(merged_data, aes(x = YEAR, y = production_difference, group = TREATMENT_kelp)) +
  geom_point(size = 3) +  # Points in skyblue color
  geom_line() +  # Lines in skyblue color
  geom_errorbar(aes(ymin = production_difference - se_difference, ymax = production_difference + se_difference), 
                width = 0.2, color = "black", size = 0.7) +  # Error bars
  labs(x = "Year",  # X-axis label
       y = expression("Kelp Habitat Fish Production" ~ (g ~ m^-2 ~ yr^-1)),  # Y-axis label
       color = "Treatment") +  # Legend title
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
        axis.title.x = element_text(vjust = -0.5, size = 18),
        axis.title.y = element_text(vjust = 1.5, size = 18),
        axis.text.y = element_text(size = 18),
        legend.position = "none");production_diff_plot

# Update production_diff_plot with annotations
production_diff_plot_annot <- production_diff_plot +
  geom_text(aes(label = sprintf("%.2f", production_difference), y = production_difference),
            color = "black", size = 4, vjust = 2.8, hjust = 1.06, check_overlap = TRUE);production_diff_plot_annot

# Save the plot
ggsave("output/figures/kelp.fish.production.png", production_diff_plot_annot, width = 20, height = 10, units = "in", dpi = 300)

@@

# Print the average
average_production_kelp # 4.359285 g/m2/yr

# Check normality and run appropriate statistical test:

# Separate data into kelp and control
kelp_data <- calc.prod %>% filter(TREATMENT == "kelp") %>% pull(production)
control_data <- calc.prod %>% filter(TREATMENT == "control") %>% pull(production)

# Shapiro-Wilk normality test
shapiro_kelp <- shapiro.test(kelp_data)
shapiro_control <- shapiro.test(control_data)

# Print results
print(shapiro_kelp)
print(shapiro_control)

# Data is not normally distributed so run Mann-Whitney U Test (Wilcoxon Rank-Sum Test)
# Perform Mann-Whitney U Test
wilcox_test_result <- wilcox.test(kelp_data, control_data)

# Print results
print(wilcox_test_result)

# Calculate mean and standard error for each treatment
kelp.prod.dat <- calc.prod %>%
  group_by(TREATMENT) %>%
  summarize(mean_production = mean(production, na.rm = TRUE),
            se_production = sqrt(sum(production^2)) / n())

# Create the plot with mean adn st error
# Create the base plot with point ranges
kelp.treatment <- ggplot(kelp.prod.dat, aes(x = TREATMENT, y = mean_production, fill = TREATMENT)) +
  geom_pointrange(aes(ymin = mean_production - se_production, ymax = mean_production + se_production), 
                  size = 1.2, color = "black", shape = 21, fill = "white") +
  labs(x = "Treatment",
       y = expression("Fish Production" ~ (g ~ m^-2 ~ yr^-1))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
        axis.title.x = element_text(vjust = -0.5, size = 18),
        axis.title.y = element_text(vjust = 1.5, size = 18),
        axis.text.y = element_text(size = 18),
        legend.position = "none"); kelp.treatment

# Calculate the maximum value for y-axis
max_y <- max(kelp.prod.dat$mean_production + kelp.prod.dat$se_production)

# Add the Wilcoxon test results as an annotation
kelp.treatment <- kelp.treatment + annotate("text", x = 0.5, y = max_y, 
                        label = "Wilcoxon test\np < 0.0001",
                        hjust = 0, vjust = 1, size = 5, color = "black");kelp.treatment

# Add significance asterisks
kelp.treatment <- kelp.treatment + geom_text(aes(x = 1.5, y = max_y - 0.2 , label = "***"), size = 6, color = "black"); kelp.treatment

# Plot all two panels together using patchwork
```{r}

# Combine the plots with patchwork
prod_combined_plot <- (production_plot | kelp.treatment) +
  plot_annotation(tag_levels = 'A') + 
  theme(plot.tag = element_text(size = 18, face = "bold"));prod_combined_plot

# Save the combined plot
ggsave("output/figures/fish_production.png", prod_combined_plot, width = 20, height = 10, units = "in", dpi = 300)
